A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2020-02-16 paper/main.tex
finalises up to evaluation of performance before sending it off to Marc

The top 15 strategies for each tournament type based on \(\bar{r}\) are given in
Table~\ref{table:top_performances}. Table~\ref{table:top_performances} includes
the top 15 strategies in noisy tournaments with \(p_n < 0.1\) and probabilistic
ending tournaments with \(p_e < 0.1\). These are included because the most common
values in the literature for the probabilities of noise and a match ending are
less than 0.1.

@@

In noisy tournaments the top strategies differ when \(p_n\) is restricted and
when it is not. The \(r\) distributions of the top 6 strategies when \(p_n\in
[0, 1]\) are given by Figure~\ref{fig:effect_of_noise}. It is evident that these
strategies were highly ranked due to the their performance in tournaments with
\(p_n>0.5\), and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly. In comparison, when considering a \(p_n < 0.1\) the
performances of the top strategies are more stable,
Figure~\ref{fig:noise_results}. The top strategies in noisy tournaments with
\(p_n < 0.1\) include DBS, Evolved FSM 16 Noise 05, Evolved ANN 5 Noise 05, PSO
Gambler 2 2 2 Noise 05 and Omega Tit For Tat~\cite{kendall2007iterated} which
are strategies specifically designed for noisy tournaments, strategies such as
BackStabber, DoubleCrosser, Evolved ANN, FSM and HMM which performed well in
standard tournaments as well, and deterministic strategies such as Spiteful Tit For
Tat~\cite{prison}, Level Punisher~\cite{Eckhart2015}, Eugine Nier~\cite{lesswrong}.

@@

In probabilistic ending tournaments it appears that the top ranks are mostly
occupied by defecting strategies and a series of strategies based on finite state
automata introduced by Daniel Ashlock and Wendy Ashlock; Fortress 3, Fortress 4
(both introduced in~\cite{Ashlock2006}), Raider~\cite{Ashlock2014} and
Solution B1~\cite{Ashlock2014}. Values of \(p_e > 0.5\) effectively turn an IPD
match to a one/two shot game, thus it is not surprise that defecting strategies
perform well. From Figure~\ref{fig:effect_of_probend}, which gives the
distributions of \(r\) for the top 6 strategies in probabilistic ending tournaments
over \(p_e\), shows that at the point of \(p_e=0.1\) defecting strategies become
the dominant strategies in their respective tournaments. In essence, what is
demonstrated in Figure~\ref{fig:effect_of_probend} is that defecting strategies
do better when the likelihood of the game ending in the next turn increased,
which is inline with the Folk Theorem~\cite{Fudenberg2009}.

@@

The top spots are not not dominated by just defecting strategies anymore when
\(p_e < 0.1\) is considered, Table~\ref{table:top_performances}. The effective
strategies are now Meta strategies which include Nice Meta Winner, NMWE Finite
Memory, NMWE Deterministic, trained strategies, Grudger~\cite{axelrodproject}
and Spiteful Tit for Tat~\cite{prison}. The Meta
strategies~\cite{axelrodproject} create a team of strategies for themselves and
play as an ensemble or some other combination of the team.
The distributions of the normalised rank in probabilistic ending tournaments,
shown in Figure~\ref{fig:probend_results}, are less variant than those of noisy
tournaments. The medians of the top 15 strategies are lower than 0.1 and the
distributions are skewed towards 0. The strategies have overall performed well
in the probabilistic ending tournaments that they participated.

In tournaments with both noise and an unspecified number of turns several of the
top ranked strategies are strategies that were highly ranked in noisy
tournaments. However, strategies from the top ranks in probabilistic ending
tournaments did not rank highly here.

The distributions of \(r\) shown in Figure~\ref{fig:noisy_probend_results} have
the largest median values compared to the top rank strategies of the other
tournament types. A subset of noisy probabilistic ending tournaments has been
considered such that \(p_e < 0.1\) and \(p_n < 0.1\). The top ranked strategies
are given in Table~\ref{table:subset_probend_noise_result} and it is shown that
the Meta strategies which performed well in noisy tournaments with \(p_n <
0.1\), perform well once again even the number of turns is not specified.
Moreover, several strategies that did well in probabilistic ending tournaments
such as Fortress 4, MEM2 and Spiteful Tit For Tat.

@@

Up till now, the performances of the \numberofstrategies strategies have been evaluated for
individual tournament types. The distributions of \(r\) for the tournament types indicate that
for probabilistic ending and standard tournaments successful strategies do exist.
For these settings, the top 15 strategies have frequently ranked in
the top spots with only a few exceptions. The behaviour of the winning strategies
in probabilistic ending tournaments vary when \(p_e\) is restricted to 0.1.
For larger values of \(p_e\) the dominant strategies are defecting strategies,
and in comparison for smaller values the dominant strategies are evolved
and strategies based on teams. Similarly, in noisy tournaments the behaviour
of the dominant strategies varied when the probability of noisy was restricted
to 0.1. In noise tournaments with \(p_n \in [0, 1]\) there is much variation
in the normalised ranks, and the strategies can always guarantee a spot in the
top ranks. However, in the restricted case the dominant strategies are strategies
specifically trained for noise tournaments.

@@

This section presented the winning strategies in a series of IPD tournaments. In
standard tournaments the top spots were dominated by complex strategies that had
been trained using reinforcement learning techniques. In noisy environments,
whether the number of turns was fixed or not, the winning strategies were
deterministic strategies designed to defect if the opponent tricked them more
than a current amount of the times that they had tricked their opponent. However,
if a value of noise strictly less than 0.1 was considered, then the successful
strategies were strategies based on the behaviour of many strategies and strategies
specifically designed for noisy tournaments. In
probabilistic ending tournaments most of the highly ranked strategies were defecting
strategies and trained finite state automata, all by the authors of~\cite{Ashlock2006, Ashlock2014}.
These strategies ranked high due to their performance in tournaments where the probability of the game ending after
each turn was bigger than 0.1. In probabilistic tournaments with \(p_e\) less than 0.1
the highly ranked strategies were trained and Meta strategies.
Finally the performance of all \numberofstrategies
strategies over the \numberofalltournaments tournaments in this manuscript was
assessed on \(\bar{r}\). The top ranked strategies were a mixture of behaviours
that did well in standard tournaments, tournaments with noise and a
few Meta strategies.

@@

% \section{Acknowledgements}

% A variety of software have been used in this work:

% \begin{itemize}
%     \item The Axelrod-Python library for IPD simulations~\cite{axelrodproject}.
%     \item The Matplotlib library for visualisation~\cite{hunter2007matplotlib}.
%     \item The Numpy library for data manipulation~\cite{walt2011numpy}.
%     \item The scikit-learn library for data analysis~\cite{scikit-learn}.
% \end{itemize}

% \appendix

% \section{A summary of parameters}\label{app:parameters}

% \begin{table}[h]
%     \begin{center}
%     \resizebox{.7\textwidth}{!}{
%     \begin{tabular}{llc}
%     \toprule
% feature & feature explanation \\
%  \midrule
% stochastic                  & If a strategy is stochastic \\
% makes use of game           & If a strategy makes used of the game information  \\
% makes use of length         & If a strategy makes used of the number of turns \\
% memory usage                & The memory size of a strategy divided by the number of turns \\
% SSE                         & A measure of how far a strategy is from extortionate behaviour \\
% $C_{\text{max}}$          & The biggest cooperating rate in the tournament \\
% $C_{\text{min}}$          & The smallest cooperating rate in the tournament \\
% $C_{\text{median}}$       & The median cooperating rate in the tournament \\
% $C_{\text{mean}}$         & The mean cooperating rate in the tournament \\
% $C_r$ / $C_{\text{max}}$    & A strategy's cooperating rate divided by the maximum \\
% $C_r$ / $C_{\text{min}}$    & A strategy's cooperating rate divided by the minimum \\
% $C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median \\
% $C_r$ / $C_{\text{mean}}$   & A strategy's cooperating rate divided by the mean \\
% $C_r$                       & The cooperating ratio of a strategy \\
% $CC$ to $C$ rate            & The probability a strategy will cooperate after a mutual cooperation \\
% $CD$ to $C$ rate            & The probability a strategy will cooperate after being betrayed by the opponent \\
% $DC$ to $C$ rate            & The probability a strategy will cooperate after betraying the opponent\\
% $DD$ to $C$ rate            & The probability a strategy will cooperate after a mutual defection\\
% $p_n$                       & The probability of a player's action being flipped at each interaction \\
% $n$                         & The number of turns \\
% $p_e$                       & The probability of a match ending in the next turn \\
% $N$                         & The number of strategies in the tournament \\
% $k$                         & The number that a given tournament is repeated \\
%     \bottomrule
%         \end{tabular}}
%     \end{center}
%     \caption{The features which are included in the performance evaluation analysis.}\label{table:parameters_summary}
% \end{table}

% \section{List of strategies}\label{app:list_of_players}

% The strategies used in this study which are from Axelrod-Python library version 3.0.0.

% \begin{multicols}{3}
% 	\begin{enumerate}
% 		\input{list_of_strategies.tex}
% 	\end{enumerate}
% \end{multicols}

% \section{Correlation coefficients}\label{app:correlations}

% A graphical representation of the correlation coefficients for the features in
% Table~\ref{table:manual_features}.

% \begin{figure}[!htbp]
%         \begin{center}
%             \includegraphics[width=.75\linewidth]{../images/standard_correlation_plot.pdf}
%         \end{center}
%         \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
%         for standard tournaments}
% \end{figure}
% \begin{figure}[!htbp]
%     \begin{center}
%         \includegraphics[width=.75\linewidth]{../images/noise_correlation_plot.pdf}
%     \end{center}
%     \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
%     for noisy tournaments}
% \end{figure}
% \begin{figure}[!htbp]
%     \begin{center}
%         \includegraphics[width=.75\linewidth]{../images/probend_correlation_plot.pdf}
%     \end{center}
%     \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
%     for probabilistic ending tournaments}
% \end{figure}
% \begin{figure}[!htbp]
%     \begin{center}
%         \includegraphics[width=.75\linewidth]{../images/probend_noise_correlation_plot.pdf}
%     \end{center}
%     \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
%     for noisy probabilistic ending tournaments}
% \end{figure}
% \begin{figure}[!htbp]
%     \begin{center}
%         \includegraphics[width=.75\linewidth]{../images/merged_correlation_plot.pdf}
%     \end{center}
%     \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
%     for data set}
% \end{figure}

>>>>

The top 15 strategies for each tournament type based on \(\bar{r}\) are given in
Table~\ref{table:top_performances}. The data collection process was design such
as that the probabilities of noise and ending of the match to vary between 0 and
1. However, the most common values for these probabilities in the literature are
values less than 0.1. Thus,
Table~\ref{table:top_performances} also includes the top 15 strategies in noisy
tournaments with \(p_n < 0.1\) and probabilistic ending tournaments with \(p_e <
0.1\).

@@

In the case of noisy tournaments with \(p_n < 0.1\) the top performed strategies
include strategies specifically designed for noisy tournaments such as DBS,
Evolved FSM 16 Noise 05, Evolved ANN 5 Noise 05, PSO Gambler 2 2 2 Noise 05 and
Omega Tit For Tat~\cite{kendall2007iterated}, strategies which performed well in
standard tournaments such as BackStabber, DoubleCrosser, Evolved ANN, FSM and
HMM and deterministic strategies such as Spiteful Tit For Tat~\cite{prison},
Level Punisher~\cite{Eckhart2015}, Eugine Nier~\cite{lesswrong}. The \(r\) distributions
of these strategies are given in Figure~\ref{fig:noise_results} where it is evident
that these strategies overall performed well in noisy tournaments.

In comparison, the top ranked strategies in noisy environments when \(p_n\in [0,
1]\) include strategies which decide their actions based on the cooperations to
defections ratio, such as ShortMem~\cite{Andre2013},
Grumpy~\cite{axelrodproject} and e~\cite{axelrodproject}, and the Retaliate
strategies which are designed to defect if the opponent has tricked them more
often than x\% of the times that they have done the same. The overall
performance of these strategies is bimodal,
Figure~\ref{fig:noise_overall_results}. This indicates that although the top
ranked strategies mainly performed well, there are several tournaments that they
ranked in the bottom half. The bimodality of the \(r\) distributions can be
explained by Figure~\ref{fig:effect_of_noise} which gives the \(r\)
distributions for the top 6 strategies over the noise probability \(p_n\).
It is evident that the highly ranked strategies
were highly ranked due to the their performance in tournaments with \(p_n>0.5\),
and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_noise.pdf}
    \caption{$r$ distributions of top 15 strategies in noisy tournaments.}\label{fig:noise_overall_results}
\end{figure}

@@

In probabilistic ending tournaments with \(p_e < 0.1\) the effective strategies
are Meta strategies which include Nice Meta Winner, NMWE Finite Memory, NMWE
Deterministic, trained strategies which performed well in standard tournaments,
Grudger~\cite{axelrodproject} and Spiteful Tit for Tat~\cite{prison}. The Meta
strategies~\cite{axelrodproject} are a set of rules that create a team of
strategies and play as an ensemble or some other combination of their team
members. The distributions of the normalised rank in the 1139 probabilistic
ending tournaments with \(p_e < 0.1\), are shown in
Figure~\ref{fig:probend_results}. The medians of the top 15 strategies are lower
than 0.1 and the distributions are skewed towards 0. The strategies have overall
performed well in the probabilistic ending tournaments that they participated.

In probabilistic ending tournaments with \(p_e \in [0, 1]\) the top ranks are
mostly occupied by defecting strategies such as Better and Better, Gradual
Killer, Hard Prober (all from~\cite{axelrodproject}), Bully (Reverse Tit For
Tat)~\cite{Nachbar1992} and Defector, and a series of strategies based on finite
state automata introduced by Daniel Ashlock and Wendy Ashlock; Fortress 3,
Fortress 4 (both introduced in~\cite{Ashlock2006}), Raider~\cite{Ashlock2014}
and Solution B1~\cite{Ashlock2014}. As stated in the Folk Theorem~\cite{Fudenberg2009}
defecting strategies do better when the likelihood of the game ending in the
next turn increased. This is demonstrated by Figure~\ref{fig:effect_of_probend}, which gives the
distributions of \(r\) for the top 6 strategies in probabilistic ending tournaments
over \(p_e\). It  shows that the 6 strategies start of with a high median rank,
however, their ranked decreased as the the probability of the game ending
increased and at the point of \(p_e = 0.1\).

@@

In tournaments with both noise and an unspecified number of turns several of the
top ranked strategies are strategies that were highly ranked in noisy
tournaments. However, strategies from the top ranks in probabilistic ending
tournaments did not rank highly here.
The distributions of \(r\) shown in Figure~\ref{fig:noisy_probend_results} have
the largest median values compared to the top rank strategies of the other
tournament types. A subset of noisy probabilistic ending tournaments has been
considered such that \(p_e < 0.1\) and \(p_n < 0.1\). The top ranked strategies
are given in Table~\ref{table:subset_probend_noise_result} and it is shown that
the Meta strategies which performed well in noisy tournaments with \(p_n <
0.1\), perform well once again even the number of turns is not specified.
Moreover, several strategies that did well in probabilistic ending tournaments
such as Fortress 4, MEM2 and Spiteful Tit For Tat.

@@

Up till now, the performances of the \numberofstrategies strategies have been
evaluated for individual tournament types. The distributions of \(r\) for the
tournament types indicate that for standard, probabilistic ending and noisy
tournaments with \(p_n <0.1\) successful strategies do exist. For these settings
the top 15 strategies frequently ranked in the top spots with only a few
exceptions. The behaviour of the winning strategies in probabilistic ending
tournaments varied when \(p_e\) was restricted to 0.1. For larger values of
\(p_e\) the dominant strategies were defecting ones and for smaller values the
dominant strategies were evolved strategies and strategies based on teams.
In noisy tournaments Similarly the behaviour
of the dominant strategies varied when the probability of noisy was restricted
to. In noise tournaments with \(p_n < 0.1\) the dominant strategies were strategies
specifically trained for noise tournaments, and the performed well in most of the
tournaments they participated. However, in noisy tournaments with \(p_n \in [0, 1]\)
the performances of the top ranked strategies were bimodal. Indicating that with
high values of noisy a successful set of strategies does not exist.

@@

This section presented the winning strategies in a series of IPD tournaments. In
standard tournaments the top spots were dominated by complex strategies that had
been trained using reinforcement learning techniques. In noisy environments where
the a noise probability strictly less than 0.1 was considered the successful
strategies were strategies based on the behaviour of many strategies and strategies
specifically designed for noisy tournaments. In
probabilistic ending tournaments most of the highly ranked strategies were defecting
strategies and trained finite state automata, all by the authors of~\cite{Ashlock2006, Ashlock2014}.
These strategies ranked high due to their performance in tournaments where the probability of the game ending after
each turn was bigger than 0.1. In probabilistic tournaments with \(p_e\) less than 0.1
the highly ranked strategies were trained and Meta strategies.
Finally the performance of all \numberofstrategies
strategies over the \numberofalltournaments tournaments in this manuscript was
assessed on \(\bar{r}\). The top ranked strategies were a mixture of behaviours
that did well in standard tournaments, tournaments with noise and a
few Meta strategies.

@@

\section{Acknowledgements}

A variety of software have been used in this work:

\begin{itemize}
    \item The Axelrod-Python library for IPD simulations~\cite{axelrodproject}.
    \item The Matplotlib library for visualisation~\cite{hunter2007matplotlib}.
    \item The Numpy library for data manipulation~\cite{walt2011numpy}.
    \item The scikit-learn library for data analysis~\cite{scikit-learn}.
\end{itemize}

\appendix

\section{A summary of parameters}\label{app:parameters}

\begin{table}[h]
    \begin{center}
    \resizebox{.7\textwidth}{!}{
    \begin{tabular}{llc}
    \toprule
feature & feature explanation \\
 \midrule
stochastic                  & If a strategy is stochastic \\
makes use of game           & If a strategy makes used of the game information  \\
makes use of length         & If a strategy makes used of the number of turns \\
memory usage                & The memory size of a strategy divided by the number of turns \\
SSE                         & A measure of how far a strategy is from extortionate behaviour \\
$C_{\text{max}}$          & The biggest cooperating rate in the tournament \\
$C_{\text{min}}$          & The smallest cooperating rate in the tournament \\
$C_{\text{median}}$       & The median cooperating rate in the tournament \\
$C_{\text{mean}}$         & The mean cooperating rate in the tournament \\
$C_r$ / $C_{\text{max}}$    & A strategy's cooperating rate divided by the maximum \\
$C_r$ / $C_{\text{min}}$    & A strategy's cooperating rate divided by the minimum \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median \\
$C_r$ / $C_{\text{mean}}$   & A strategy's cooperating rate divided by the mean \\
$C_r$                       & The cooperating ratio of a strategy \\
$CC$ to $C$ rate            & The probability a strategy will cooperate after a mutual cooperation \\
$CD$ to $C$ rate            & The probability a strategy will cooperate after being betrayed by the opponent \\
$DC$ to $C$ rate            & The probability a strategy will cooperate after betraying the opponent\\
$DD$ to $C$ rate            & The probability a strategy will cooperate after a mutual defection\\
$p_n$                       & The probability of a player's action being flipped at each interaction \\
$n$                         & The number of turns \\
$p_e$                       & The probability of a match ending in the next turn \\
$N$                         & The number of strategies in the tournament \\
$k$                         & The number that a given tournament is repeated \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The features which are included in the performance evaluation analysis.}\label{table:parameters_summary}
\end{table}

\section{List of strategies}\label{app:list_of_players}

The strategies used in this study which are from Axelrod-Python library version 3.0.0.

\begin{multicols}{3}
	\begin{enumerate}
		\input{list_of_strategies.tex}
	\end{enumerate}
\end{multicols}

\section{Correlation coefficients}\label{app:correlations}

A graphical representation of the correlation coefficients for the features in
Table~\ref{table:manual_features}.

\begin{figure}[!htbp]
        \begin{center}
            \includegraphics[width=.75\linewidth]{../images/standard_correlation_plot.pdf}
        \end{center}
        \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
        for standard tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/noise_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
    for noisy tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/probend_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
    for probabilistic ending tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/probend_noise_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
    for noisy probabilistic ending tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/merged_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of features in Table~\ref{table:manual_features}
    for data set}
\end{figure}
